Experiences in teaching with Jupyter with nbgrader

### 2019-09-23 paper/paper.tex
Update the adresses
material from my Chorzów presentation added

\institute{Marcin Kostur \at
              first address \\
              Tel.: +123-45-678910\\
              Fax: +123-45-678910\\
              \email{marcin.kostur@us.edu.pl}           %  \\
%             \emph{Present address:} of F. Author  %  if needed
           \and
           Marina Appiou-Nikiforou \at
              European University Cyprus \\
              \email{M.Nikiforou@euc.ac.cy}
	   \and
	   Andreas Papadopoulos \at
              European University Cyprus \\
              \email{M.Nikiforou@euc.ac.cy}
	   \and
	   Gert-Ludwig Ingold \at
	      Institut f{\"u}r Physik, Universit{\"a}t Augsburg,
	      Universit{\"a}tsstra{\ss}e 1, 86135 Augsburg, Germany\\
	      \email{gert.ingold@physik.uni-augsburg.de}
}

@@

>>>>

\institute{Marcin Kostur \at
              first address \\
              Tel.: +123-45-678910\\
              Fax: +123-45-678910\\
              \email{marcin.kostur@us.edu.pl}           %  \\
%             \emph{Present address:} of F. Author  %  if needed
           \and
           Marina Appiou-Nikiforou \at
              European University Cyprus, 
	      6 Diogenous Street, 1516 Nicosia, Cyprus\\
              \email{M.Nikiforou@euc.ac.cy}
	   \and
	   Andreas Papadopoulos \at
              European University Cyprus, 
	      6 Diogenous Street, 1516 Nicosia, Cyprus \\
              \email{apapado89@gmail.com}
	     % \emph{Present address:}
	   \and
	   Gert-Ludwig Ingold \at
	      Institut f{\"u}r Physik, Universit{\"a}t Augsburg,
	      Universit{\"a}tsstra{\ss}e 1, 86135 Augsburg, Germany\\
	      \email{gert.ingold@physik.uni-augsburg.de}
}

@@

\section{Jupyter and Jupyterhub}

pros and cons of using the Jupyter notebook
\begin{itemize}
 \item notebook allows to guide students through a problem set
 \item problems related to out-of-order execution
 \item \textit{but} students might think that programming in Python
       necessarily implies working with a Jupyter notebook,
       they should know about IDEs (spyder, \dots) and
       editors (vim, \dots)
\end{itemize}

pros and cons of using Jupyterhub
\begin{itemize}
 \item easily accessible interface to problem sets
 \item no need to install Python on local computer
 \item consistent working environment for all students
 \item \textit{but} may be beginners should gather experience with Python
       on their own computer
\end{itemize}

\section{\texttt{nbgrader}}

\subsection{Functions}

In \texttt{nbgrader}, the solutions should be implemented in terms of a
function, so that this function can be tested, typically for different
parameters. As a consequence, the concept of functions needs to be introduced
early on in a course. However, the use of functions will be close to the
concept of a mathematical function in the sense that there will be input values
and output values. In the tests used to grade the answer, one wants to vary the
input values and the output value is the object which is to be tested against
an expectation. Therefore, situations like the absence of arguments or return
values as well as default arguments and keyword arguments representing more
advanced concepts, are not required to work with \texttt{nbgrader}.

It is good programming practice to logically structure code into functions of a
not too large length. Students with \texttt{nbgrader} thus get used to this
practice early on, in particular if functions are used to build up the final
solution step by step. The drawback however is that this structure is provided
to them by the teacher. Students are not actively structuring their code by
themselves and thus do not really exercise this practise.

In Python, the use of docstrings is highly recommended as way to document
the behavior of a function. For the use with \texttt{nbgrader}, teachers are
naturally using docstring because they need to specify the purpose of the
function and the meaning of the input and output values. Students will thus
soon consider docstring as standard part of a function. However, they do not
actively practice the habit of annotating a function with a docstring.

The version of a function released to the students by \texttt{nbgrader} by
default raises a \texttt{NotImplementedError}. This statement ensures the
correct working of the function during grading even if no solution is provided.
Students need to remove the statement and replace it by their solution. It was
found both in the German as well as the Polish group of students that this need
was somewhat difficult to convey to them. This difficulty likely arises from
the fact that students have not been taught the meaning of the \texttt{raise}
statement. They therefore need to be especially alerted to what they need to do
with this statement and internalize the removal of the \texttt{raise} statement
as a recipe.

\subsection{Grading with tests}

The main idea for autograding with \texttt{nbgrader} consists in the teacher
providing one or several tests while the student needs to develop code which
satisfies theses tests. This strategy is typical for test-driven development,
a component of agile programming where tests are written before the code.
The use of \texttt{nbgrader} thus offers an opportunity to discuss the role
of testing and the concept of test-driven development. However, it certainly
depends on the programming background of the specific students whether such
remarks are fruitful. While one should not expect that students get into the
habit of test-driven development right away, the concept might stick in their
mind and be activated at a later time.

The experience in both, the German and the Polish groups was that tests in
\texttt{nbgrader} problem sets should not only pertain to the specific
functionality of a solution to be developed by the student. It turned out
that mistakes like not returning a result or returning a result of an incorrect
type do not only occur when students start working with \texttt{nbgrader} but
also in later stages. It has therefore proven useful to separate tests into
two groups. The first groups of tests check for the existence of a result
and for the correct type. These are prerequisites for the second group of results
which test the functionality of the proposed solution for correctness.

The first group of tests should mainly be considered as an opportunity to
provide feedback to the students. They could be alerted to a missing return
value by the test

\noindent\texttt{assert result is not None, 'Does your function return
a result?'}

Quite generally, tests in \texttt{nbgrader} should not only be considered as
tool for grading but also as an opportunity to give helpful feedback to the
students while working on the problem set. Providing an expressive error
message will be helpful for many students who are not yet capable to understand
or correctly interpret the standard traceback of a Python exception.

A problem very frequently encountered with students results from detailed 
tests with clear error messages. Students were found to rely in their assessment
of correctness of their solution almost entirely on the results of the tests.
In other words, they typically do not develop a critical attitude towards their
code and do not see the need of doing their own tests. One could imagine to
have several blocks of tests to demonstrate that the passing of a number of
tests does not necessarily need that the code is correct. Setting up such a
test suite however relies on a good knowledge of typical mistakes which can
be exploited.

\subsection{Addressing heterogeneity}

It is not uncommon that in programming courses one has to address an audience
with a wide range of programming capabilities. Some students have no
programming experience whatsoever and might even show a code writing block
while others have an extensive programming experience, often in a different
programming language. For students with little or no programming experience,
more complex exercise problems need to be decomposed into a number steps while
experienced students might tackle the problems directly without any additional
help. The latter students often are proficient in a language different from
Python. A typical indicator are loops over indiced instead of objects contained
in a list. Here, the feedback mechanism of \texttt{nbgrader} comes handy, in
particular for \texttt{nbgrader} version 0.6+.

At the same time, the solution of the problem should provide a source of
satisfaction for all students. When teaching programming to science students, a
way could be to appeal to the respective scientific background. However, such
problem sets are often experienced as an unnecessary mental burden.

In exercise work with \texttt{nbgrader}, we have experimented with allowing for
individual learning paths by providing one main Jupyter notebook and typically
a few subnotebooks. The main notebook provides the problem description and the
notebook cells where the final solution should be developed. The subnotebooks
focus on one aspect of the solution so that students can work towards a complete
solution step by step.

To illustrate our approach by way of an example, we consider a problem set
where a graphical representation of a Julia set shall be generated. Students
need to repetitively carry out an interation $z\rightarrow z^2+c$ where $c$ is
a given complex constant and the iteration starts from complex values $z_0$
taken on a grid in the complex plane. The Julia set distinguishes between the
initial conditions for which $z$ converges from those where $z$ remains bounded.
In practice, this distinction can be achieved by checking whether an appropriately
chosen bound is crossed within a given maximum number of iterations. A beautiful
graphical representation can be obtained by converting the number of iterations
actually needed to cross the bound into a color.

This problem presents a number of difficulties to the students. From an
algorithmic point of view, the iteration needs to be implemented and stopped
once either the bound is crossed or the maximum number of iterations has been
reached. On a more conceptual level, students were often found to encounter
problems in imagining how a computer can represent colors. A subnotebook
explaining the concept of a colormap and allowing students to explore the
translation of a number into a color representation turns out to be helpful.
Here, the rich output of the Jupyter notebook is used by providing a SVG
representation of a square filled with the respective color. Care needs to
be taken that the students are not irritated by the code needed to provide
the infrastructure for generating the colored square. The opportunity of
playing with the conversion of a number into a visual representation of a
color usually turns out to be very helpful. At this point, a background of
the RGB or HSV representation of colors is not necessarily required but
can be offered to interested students in a separate subnotebook.

Another conceptual obstacle for beginners is the idea of pixels in an image
and how to walk along a two-dimensional grid or even a one-dimensional grid
within a program. This aspect thus deserves a separate subnotebook and proceeding
in steps from one to two dimensions proved to be useful. The main reason for
addressing the one-dimensional case first is due to the observation that if
students are asked to produce a given number of equally spaced points between
two values, they typically confuse the number of points with the number of
intervalls between the points. This issue needs to be first clarified in one
dimension before the two-dimensional case can be addressed.

Structuring a problem into a number of steps with their own notebooks allows to
offer individual learning paths. In the supervised exercise sessions it turned
out that the vast majority of students tends to follow the step-by-step
learning path which led most directly to the solution. In other words, students
would have needed very strong encouragement to try to develop the solution
completely on their own. On the other hand, students typically did not go through
the additional material provided on the RGB and HSV representation.

On a more technical level, it could be observed that the fact that opening a
new notebook would lead to an extra tab in the browser could become quite
confusing for the students unless the take care to close tabs not needed
anymore. In particular, it happened that the students opened notebooks more
than once and then getting confused in which notebook they were actually
working.

Using \texttt{nbgrader} in a scenario as described above, an additional
difficulty arises in the grading. Students working on subnotebooks should earn
the respective points even if they are able to only solve some of the steps
successfully. On the other hand, a student who solves the problem without going
through all the notebooks should not be punished. The \texttt{nbgrader}
database holds all the points obtained for the various notebooks, from which a
final number of points can be determined. This is however a functionality not
directly provided by \texttt{nbgrader}. If points are assigned appropriately to
the individual tasks, one can determine the number of points earned as the
minimum of the total sum of points earned and the maximum number of points
attributed to the problem set.

\section{Conclusion}
